Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-08 text/manuscript-full-KS.qmd
revisions dallas rq2

We explore socioeconomic demographics -- median income, percent living below poverty, and educational attainment -- within and across Geomarkets over time in @fig-chicago-rq1-ses. The primary takeaways are the City of Chicago -- and to a somewhat lesser degree th South and Southwest Suburbs - stands out as having lower median income, higher poverty rates, and lower BA attainment than other Geomarkets. By contrast, the North Shore stands out as being particularly affluent and highly educated. Overall, median income both within and across Geomarkets in the Chicago metropolitan area remained relatively stable across the three census periods. For instance, in 1980, median household income across the Chicago metropolitan area ranged from a high of \$168,000 in the North Shore Geomarket to a low of \$68,000 in the City of Chicago Geomarket. However, median incomes in all other Geomarkets were closer to the upper thresholds of affluence, such as \$131,000 in Evanston and Skokie, \$120,000 in the Northwest Suburbs, \$111,000 in the Western Suburbs, \$109,000 in Chain of Lakes, and \$99,000 in the South and Southwest Suburbs. While overall income patterns across Geomarkets remained consistent through 2020, a median income increase for the City of Chicago Geomarket (\$80,000) and decrease in the South and Southwest Suburbs Geomarket (\$91,000) narrowed the income gap between the lowest and second-lowest Geomarkets to just \$11,000. @fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago and South and Southwest Suburbabs had the lowest percentage of residents with a bachelor's degree or higher (14%). However, by 2020, the City of Chicago experienced the largest percentage point increase in college degree attainment, reaching 34%.

@@

Socioeconomic demographics of median income, percent living below poverty, and educational attainment over time are presented in @fig-dallas-rq1-ses for the Dallas-Fort Worth metropolitan area. Overall, median income both within and across Geomarkets remained relatively stable across the three census periods. <!-- Between 1980 to 2000, affluence levels across Geomarkets were consistent, ranging from a low of \$82,000 in City of Fort Worth to a high of \$135,00 in the Collin & Rockwall Counties Geomarket. However, the substantial increase in median income for the Collin & Rockwall Counties Geomarket during this time period plateaued by 2020, reshuffling relatively affluence levels among Geomarkets. By 2020, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$133,000, followed by West of Dallas/Fort Worth Metroplex (\$116,000), and Dallas County (\$90,000). In contrast, median household income for City of Dallas and City of Fort Worth remained relatively stable through 2020. As a result, --> In 2020, both City of Dallas and City of Fort Worth continued to rank at the lowest levels of affluence within the metropolitan area at \$81,000 and \$86,000 median household incomes, respectively. By contrast, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$129,000, followed by West of Dallas/Fort Worth Metroplex (\$120,000). @fig-dallas-rq1-ses also illustrates changes in the percentage of residents living in poverty and educational attainment within Geomarkets over time, revealing similar dynamics across Geomarkets to those observed in median household income. Together, these trends underscore the growingly racial/ethnic diverse but widening economic inequality within the evolving Geomarket landscape of the Dallas-Fort Worth metroplex.

@@

@fig-dallas-rq2-race presents the racial/ethnic composition of purchased student profiles within Geomarkets in the Dallas-Forthworth area, revealing a comparatively more distributed pattern of racial/ethnic diversity across Geomarkets unlike Chicago's concentrated diversity within a single Geomarket. This distribution is most notable at the lowest PSAT score threshold. For instance, at the 1070-1180 PSAT score range, the central suburban Geomarkets of Dallas County (36% White, 17% Asian, 15% Black, 28% Hispanic) and Irving, Arlington, and Grand Prairie (34% White, 13% Asian, 15% Black, 34% Hispanic) have the greatest proportions of non-White prospects, which is largely attributed to their greater shares of Asian and Black prospects. While the urban central Geomarkets of City of Dallas and City of Fortworth are the most racially/ethnically diverse in the larger metropolitan area (see @fig-dallas-rq1-race), their prospect pools in @fig-dallas-rq2-race includes larger proportions of Hispanic prospects but lower proportions of Asian and Black students. The Collin & Rockwall Counties (57% White, 7% Black, 17% Hispanic, 14% Asian) and West of Dallas/Forth Worth Metroplex Geomarkets (61% White, 7% Black, 19% Hispanic, 7% Asian) both yield the largest but predominantly White prospect pools. Prospect pools within Geomarkets from the middle range PSAT score (1190-1260) to the high PSAT score range (1270-1520) become disproportionately more Asian, while the representation of Hispanic and Black prospects decline substantially.

@fig-dallas-rq2-midSAT-race presents the race/ethnicity of all prospects targeted in the middle PSAT score range across Geomarket contributions. The bars for "All (race known)" show that Collin & Rockwall (37%) and West of Dallas/Forth Worth (23%) contribute many more students than the other four Geomarkets. Disparities in Geomarket contributions were evident but differed for Asian, Black, and Hispanic students. For instance, The Dallas County (20% versus 13% for all prospects) and Colin and Rockwall Counties (47% versus 37% for all prospects) Geomarkets contributed a disproportionate larger share of Asian students, whereas nearly all other Geomarkets contributed disproportionately smaller shares.

@@

Representational patterns across Geomarkets are reversed for Black and Hispanic prospects in @fig-dallas-rq2-midSAT-race. All Geomarkets besides Collin and Rockwall Counties and West of Dallas/Fort Worth Metroplex contributed equal or larger proportions of Black and Hispanic prospects relative to their contributions to the overall prospect pool in the Dallas-Fort Woth metropolitan area. However, the City of Dallas Geomarket contributed the largest disproportionate share of Hispanic prospects (19% versus 11% for all prospects), whereas the Dallas Counties Geomarket contributed the largest disproportionate share of Black prospects (21% versus 13% for all prospects).

Contributions of each Geomarket to the pool of prospects based on first-generation college student status from the Dallas-Fort Worth metropolitan area are presented in @fig-dallas-rq2-midSAT-ses [THE FIGURE IN THE PAPER IS WRONG]. Only 303 prospects of the 3,928 total in this middle PSAT score range pool were first-generation college students whose parents did not attend college. Again, all Geomarkets besides Collin and Rockwall Counties (14% versus 36% for all prospects) and West of Dallas/Fort Worth Metroplex (15% versus 24% for all prospects) contributed equal or larger proportions of first-generation college students (no college) relative to their contributions to the overall prospect pool. Disproportioantely larger contributions range from six percentage points by the Dallas Counties Geomarket (19% versus 13% for all prospects) to 11 percentage points by the City of Dallas Geomarket (22% versus 11% for all prospects). Contributions become more proportional for the pool of first-generation college students whose parents attended but did not complete their degree. Proportional balances across Geomarkets are also evident for the 3,069 prospects that are not first-generation college students, with the exception of Collin and Rockwall Counties Geomarket contributing the largest disproportionate share of not first-generation college students (40%) relative to the contributions to the overall prospect pool (36%).

Lastly, @fig-dallas-rq2-midSAT-combo presents Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity, which we view as the most interesting results for the Dallas-Fort Worth metropolitan area. Among all White prospects (n=2,094) whose contact information was purchased in this middle PSAT score range, approximately 3% were first-generation college students whose parents did not attend college, 11% were first-generation college students whose parents attended some college but did not complete their degree, and 86% were not first-generation college students. @fig-dallas-rq2-midSAT-combo shows nearly all Geomarkets contribute a nearly proportionate share of prospects by first-generation college students relative to the overall pool, the only modest exception being the Irving, Arlington, & Grand Prairie Geomarket.

By contrast, for non-White groups, a larger share of prospects were first-generation students. Furthermore, first generation status differed substantially across Geomarkets, especially for Hispanic and Asian students. For Hispanic students, more than 50% of prospects from the City of Dallas, City of Forth Worth, Arlington & Grand Prairie, and Dallas County are first-generation college students. By contrast, for the more affluent Collin & Rockwall Geomarket and the West of Dallas/Fort Worth Metroplex Geomarkets, less than 30% of prospects were first-generation. For Asian prospects, City of Fort Worth and Irving, Arlington, & Grand Prairie Geomarkets contributed disproportionately larger shares of first-generation college students across both parents with no college and some college. By contrast, the more affluent Collin & Rockwall sent more Asian prospects (285) than any other Geomarket, but about 87% of these prospects were not first-generation college students.

@@

```{r fig-chicago-rq2-midSAT-combo, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarket Contribution to Purchased Student profiles by First-Generation Status, Middle-Range SAT orders", out.width = "100%"}
#
knitr::include_graphics('./../results/graphs/rq2b/rq2b_chicago_order_392834_488035_race_by_firstgen.png')
#knitr::include_graphics('./../assets/images/rq2_chicago_midSAT_combo.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

@@

```{r fig-dallas-rq2-race, echo = FALSE, fig.align = 'center', fig.cap = "Racial/Ethnic Composition of Purchased Student Profiles by Geomarket, Dallas Area", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq2-dallas-race.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-dallas-rq2-midSAT-race, echo = FALSE, fig.align = 'center', fig.cap = "Dallas Geomarket Contribution to Purchased Student profiles by Racial/ethnic group, Middle-Range SAT orders", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq2_dallas_midSAT_race.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)

@@

```{r fig-dallas-rq2-midSAT-ses, echo = FALSE, fig.align = 'center', fig.cap = "Dallas Geomarket Contribution to Purchased Student profiles by First-Generation Status, Middle-Range SAT orders", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq2_chicago_midSAT_ses.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-dallas-rq2-midSAT-combo, echo = FALSE, fig.align = 'center', fig.cap = "Dallas Geomarket Contribution to Purchased Student profiles by First-Generation Status, Middle-Range SAT orders", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq2_dallas_midSAT_combo.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

>>>>

We explore socioeconomic demographics -- median income, percent living below poverty, and educational attainment -- within and across Geomarkets over time in @fig-chicago-rq1-ses. The primary takeaways are the City of Chicago -- and to a somewhat lesser degree th South and Southwest Suburbs - stands out as having lower median income, higher poverty rates, and lower BA attainment than other Geomarkets. By contrast, the North Shore stands out as being particularly affluent and highly educated. Overall, median income both within and across Geomarkets in the Chicago metropolitan area remained relatively stable across the three census periods. For instance, in 1980, median household income across the Chicago metropolitan area ranged from a high of \$165,000 in the North Shore Geomarket to a low of \$68,000 in the City of Chicago Geomarket. However, median incomes in all other Geomarkets were closer to the upper thresholds of affluence, such as \$125,000 in Evanston and Skokie, \$117,000 in the Northwest Suburbs, \$108,000 in the Western Suburbs, \$104,000 in Chain of Lakes, and \$99,000 in the South and Southwest Suburbs. While overall income patterns across Geomarkets remained consistent through 2020, a slight median income increase for the City of Chicago Geomarket (\$71,000) and decrease in the South and Southwest Suburbs Geomarket (\$84,000) narrowed the income gap between the lowest and second-lowest Geomarkets to just \$13,000. @fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago and South and Southwest Suburbs had the lowest percentage of residents with a bachelor's degree or higher (14%). However, by 2020, the City of Chicago experienced the largest percentage point increase in college degree attainment, reaching 34%.

@@

Socioeconomic demographics of median income, percent living below poverty, and educational attainment over time are presented in @fig-dallas-rq1-ses for the Dallas-Fort Worth metropolitan area. Overall, median income both within and across Geomarkets remained relatively stable across the three census periods. <!-- Between 1980 to 2000, affluence levels across Geomarkets were consistent, ranging from a low of \$82,000 in City of Fort Worth to a high of \$135,00 in the Collin & Rockwall Counties Geomarket. However, the substantial increase in median income for the Collin & Rockwall Counties Geomarket during this time period plateaued by 2020, reshuffling relatively affluence levels among Geomarkets. By 2020, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$133,000, followed by West of Dallas/Fort Worth Metroplex (\$116,000), and Dallas County (\$90,000). In contrast, median household income for City of Dallas and City of Fort Worth remained relatively stable through 2020. As a result, --> In 2020, both City of Dallas and City of Fort Worth continued to rank at the lowest levels of affluence within the metropolitan area at \$68,000 and \$74,000 median household incomes, respectively. By contrast, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$124,000, followed by West of Dallas/Fort Worth Metroplex (\$104,000). @fig-dallas-rq1-ses also illustrates changes in the percentage of residents living in poverty and educational attainment within Geomarkets over time, revealing similar dynamics across Geomarkets to those observed in median household income. Together, these trends underscore the growingly racial/ethnic diverse but widening economic inequality within the evolving Geomarket landscape of the Dallas-Fort Worth metroplex.

@@

@fig-dallas-rq2-race presents the racial/ethnic composition of purchased student profiles within Geomarkets in the Dallas-Forthworth area, revealing a comparatively more distributed pattern of racial/ethnic diversity across Geomarkets unlike Chicago's concentrated diversity within a single Geomarket. This distribution is most notable at the lowest PSAT score threshold. For instance, at the 1070-1180 PSAT score range, the central suburban Geomarkets of Dallas County (37% White, 17% Asian, 15% Black, 27% Hispanic) and Irving, Arlington, and Grand Prairie (34% White, 13% Asian, 15% Black, 33% Hispanic) have the greatest proportions of non-White prospects, which is largely attributed to their greater shares of Asian and Black prospects. While the urban central Geomarkets of City of Dallas and City of Fortworth are the most racially/ethnically diverse in the larger metropolitan area (see @fig-dallas-rq1-race), their prospect pools in @fig-dallas-rq2-race includes larger proportions of Hispanic prospects but lower proportions of Asian and Black students. The Collin & Rockwall Counties (58% White, 7% Black, 17% Hispanic, 13% Asian) and West of Dallas/Forth Worth Metroplex Geomarkets (60% White, 7% Black, 18% Hispanic, 9% Asian) both yield the largest but predominantly White prospect pools. Prospect pools within Geomarkets from the middle range PSAT score (1190-1260) to the high PSAT score range (1270-1520) become disproportionately more Asian, while the representation of Hispanic and Black prospects decline substantially.

@fig-dallas-rq2-midSAT-race presents the race/ethnicity of all prospects targeted in the middle PSAT score range across Geomarket contributions. The bars for "All (race known)" show that Collin & Rockwall (30%) and West of Dallas/Forth Worth (27%) contribute many more students than the other four Geomarkets. Disparities in Geomarket contributions were evident but differed for Asian, Black, and Hispanic students. For instance, The Dallas County (19% versus 12% for all prospects) and Colin and Rockwall Counties (39% versus 30% for all prospects) Geomarkets contributed a disproportionate larger share of Asian students, whereas nearly all other Geomarkets contributed disproportionately smaller shares.

@@

Representational patterns across Geomarkets are reversed for Black and Hispanic prospects in @fig-dallas-rq2-midSAT-race. All Geomarkets besides Collin and Rockwall Counties and West of Dallas/Fort Worth Metroplex contributed nearly equal or larger proportions of Black and Hispanic prospects relative to their contributions to the overall prospect pool in the Dallas-Fort Woth metropolitan area. However, the City of Dallas Geomarket contributed the largest disproportionate share of Hispanic prospects (19% versus 12% for all prospects), whereas the Irving, Arlington, and Grand Prairie Geomarket contributed the largest disproportionate share of Black prospects (15% versus 7% for all prospects).

Contributions of each Geomarket to the pool of prospects based on first-generation college student status from the Dallas-Fort Worth metropolitan area are presented in @fig-dallas-rq2-midSAT-ses. Only 303 prospects of the 3,926 total in this middle PSAT score range pool were first-generation college students whose parents did not attend college. Again, all Geomarkets besides Collin and Rockwall Counties (12% versus 30% for all prospects) and West of Dallas/Fort Worth Metroplex (16% versus 27% for all prospects) contributed equal or larger proportions of first-generation college students (no college) relative to their contributions to the overall prospect pool. Disproportionately larger contributions range from five percentage points by the City of Fort Worth (16% versus 11% for all prospects) to 10 percentage points by the City of Dallas Geomarket (22% versus 12% for all prospects). Contributions become more proportional for the pool of first-generation college students whose parents attended but did not complete their degree. Proportional balances across Geomarkets are also evident for the 3,067 prospects that are not first-generation college students. <!--#  , with the exception of Collin and Rockwall Counties Geomarket contributing the largest disproportionate share of not first-generation college students (33%) relative to the contributions to the overall prospect pool (30%).-->

Lastly, @fig-dallas-rq2-midSAT-combo presents Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity, which we view as the most interesting results for the Dallas-Fort Worth metropolitan area. Among all White prospects (n=2,093) whose contact information was purchased in this middle PSAT score range, approximately 3% were first-generation college students whose parents did not attend college, 11% were first-generation college students whose parents attended some college but did not complete their degree, and 86% were not first-generation college students. @fig-dallas-rq2-midSAT-combo shows almost all Geomarkets contribute a nearly proportionate share of prospects by first-generation college students relative to the overall pool, the only modest exception being the Irving, Arlington, & Grand Prairie Geomarket.

By contrast, for non-White groups, a larger share of prospects were first-generation students. Furthermore, first generation status differed substantially across Geomarkets, especially for Hispanic and Asian students. For Hispanic students, more than 50% of prospects from the City of Dallas, City of Forth Worth, Arlington & Grand Prairie, and Dallas County are first-generation college students. By contrast, for the more affluent Collin & Rockwall Geomarket and the West of Dallas/Fort Worth Metroplex Geomarkets, less than 30% of prospects were first-generation. For Asian prospects, City of Fort Worth and Irving, Arlington, & Grand Prairie Geomarkets contributed disproportionately larger shares of first-generation college students across both parents with no college and some college. By contrast, the more affluent Collin & Rockwall sent more Asian prospects (234) than any other Geomarket, but about 89% of these prospects were not first-generation college students.

@@

```{r fig-chicago-rq2-midSAT-combo, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarket Contribution to Purchased Student profiles by First-Generation Status, Middle-Range SAT orders", out.width = "100%"}

knitr::include_graphics('./../results/graphs/rq2b/rq2b_chicago_order_392834_488035_race_by_firstgen.png')
#knitr::include_graphics('./../assets/images/rq2_chicago_midSAT_combo.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

@@

```{r fig-dallas-rq2-race, echo = FALSE, fig.align = 'center', fig.cap = "Racial/Ethnic Composition of Purchased Student Profiles by Geomarket, Dallas Area", out.width = "100%"}
knitr::include_graphics('./../results/graphs/rq2/rq2_dallas_race_row_plot.png')
#knitr::include_graphics('./../assets/images/rq2-dallas-race.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-dallas-rq2-midSAT-race, echo = FALSE, fig.align = 'center', fig.cap = "Dallas Geomarket Contribution to Purchased Student profiles by Racial/ethnic group, Middle-Range SAT orders", out.width = "100%"}
knitr::include_graphics('./../results/graphs/rq2/rq2_dallas_race_col_plot_448427.png')
#knitr::include_graphics('./../assets/images/rq2_dallas_midSAT_race.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)

@@

```{r fig-dallas-rq2-midSAT-ses, echo = FALSE, fig.align = 'center', fig.cap = "Dallas Geomarket Contribution to Purchased Student profiles by First-Generation Status, Middle-Range SAT orders", out.width = "100%"}
knitr::include_graphics('./../results/graphs/rq2/rq2_dallas_firstgen_col_plot_448427.png')

#knitr::include_graphics('./../assets/images/rq2_chicago_midSAT_ses.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-dallas-rq2-midSAT-combo, echo = FALSE, fig.align = 'center', fig.cap = "Dallas Geomarket Contribution to Purchased Student profiles by First-Generation Status, Middle-Range SAT orders", out.width = "100%"}

knitr::include_graphics('./../results/graphs/rq2b/rq2b_dallas_order_448427_race_by_firstgen.png')
#knitr::include_graphics('./../assets/images/rq2_dallas_midSAT_combo.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```
